---
title: "Manuscript"
number-sections: true
format:
  html:
    toc: true
    toc-depth: 4
editor: source
editor_options: 
  chunk_output_type: console
#bibliography: references.bib
#csl: the-american-naturalist.csl
---

```{r include = FALSE, eval = FALSE}
# The code that adds the .nojekyll file to the GitHub repository:
file <- ".nojekyll"
file.create(file)
gert::git_add(file)
gert::git_commit("Adding the .nojekyll file")
gert::git_push()
rm(file)
```

```{r include = FALSE, eval = FALSE}
run_all <- function() {
  knitr::purl("manuscript.qmd", "tmp.R", documentation = FALSE)
  source("tmp.R")
  file.remove("tmp.R")  
}

rerun_all <- function() {
  rm(list  = grep("run_all", ls(envir = .GlobalEnv), value = TRUE, invert = TRUE),
     envir = .GlobalEnv)
  run_all()
}
```

```{r include = FALSE, purl = FALSE}
setHook("plot.new", NULL, action = "replace")

par3 <- function(...) par(..., mgp = c(1.5, .5, 0), bty = "n")

knitr::knit_hooks$set(
  margin = function(before, options, envir) {
    if (before) par3(plt = c(.105, .97, .15, .95)) else NULL
  })

figw <- 7  # default is 7 and maximum possible is 8.3
figh <- 5  # default is 5
eps <- .8

knitr::opts_chunk$set(margin     = TRUE,
                      fig.retina = 2,
                      fig.align  = "center",
                      fig.width  = eps * figw,
                      fig.height = eps * figh) 
```


## Constants

Path to data:

```{r}
user <- Sys.getenv("USER")
if (user == "MarcChoisy") {
  path2data <- paste0("~/Library/CloudStorage/OneDrive-SharedLibraries-OxfordUniversi",
                      "tyClinicalResearchUnit/OUCRU Modelling - Documents/HFMD HCMC 2",
                      "023/data/cleaned/")
} else if (user == "marc") {
  path2data <- paste0("~/Library/CloudStorage/OneDrive-SharedLibraries-OxfordUniversi",
                      "tyClinicalResearchUnit/OUCRU Modelling - HFMD HCMC 2023/data/c",
                      "leaned/")
} else {
  path2data <- paste0("D:/OUCRU-Onedrive/OneDrive - Oxford University Clinical Resear",
                      "ch Unit/Data/HFMD/cleaned/")
}
```

Files names:

```{r}
cases_file <- "hfmd_hcdc.rds"
sero_file <- "sero_EV-A71_1222-1223.rds"
viro_file <- "virology_cleaned.rds"
```


## Packages

Required packages:

```{r}
required <- c("magrittr", "lubridate", "stringi", "stringr", "purrr", "tidyr",
              "ISOweek", "mclust", "mgcv", "dplyr")
```

Installing those that are not installed yet:

```{r}
to_install <- required[! required %in% installed.packages()[, "Package"]]
if (length(to_install)) install.packages(to_install)
```

Loading the packages for interactive use:

```{r warning = FALSE, message = FALSE}
invisible(sapply(required, library, character.only = TRUE))
```


## Functions

A tuning of `readRDS()`:

```{r}
readRDS2 <- function(x) readRDS(paste0(path2data, x))
```

Some tunings of the `plot()` function:

```{r}
plotd <- function(...) plot(..., lwd = 2, main = NA)
plotl <- function(...) plot(..., type = "l", lwd = 2)
```

A tuning of `lines()`:

```{r}
lines2 <- function(...) lines(..., lwd = 2)
```

A tuning of `density()`:

```{r}
density2 <- function(...) density(..., n = 2048)
```


## Data

Serological data:

```{r}
df_sero <- readRDS2(sero_file)
```

Virological data:

```{r}
df_viro <- readRDS2(viro_file)
```

Case report data:

```{r}
df23 <- cases_file |>
  readRDS2() |> 
  select(dob, admission_date, district, inout, severity, medi_cen) |> 
  filter(year(admission_date) == 2023) |> 
  filter_out(is.na(dob)) |> 
  unique() |> 
  mutate(across(severity, ~ case_when(is.na(.x) ~ "Undefined",
                                      severity %in% c("3", "4") ~ "3 or 4",
                                      .default = severity) |> as.factor()),
         across(medi_cen, ~ tolower(stri_trans_general(.x, "Latin-ASCII"))),
         across(medi_cen,
                ~ case_when(.x == "benh vien nhi dong 1" ~ "Children's hospital 1",
                            .x == "benh vien benh nhiet doi tphcm" ~
                              "Hospital of Tropical Diseases",
                            .x == "benh vien nhi dong 2" ~ "Children's hospital 2",
                            .x == "benh vien nhi dong thanh pho" ~
                              "City Children's hospital", .default = .x)),
         age1     = interval(dob, admission_date) / years(1),
         time     = as.numeric(admission_date),
         adm_week = as.Date(floor_date(admission_date, "week")),
         age      = round(age1))
```


## Weekly incidence

Calculate incidence from `start` by steps of `by`:

```{r}
incidence_from <- function(start, by) {
  df23 |> 
    mutate(week = cut(time, with(df23, seq(start, max(time), by)),
                      include.lowest = TRUE)) |> 
    group_by(week) |> 
    tally() |> 
    mutate(across(week, as.character),
           across(week, ~ .x |>
                    str_remove("\\[") |> str_remove("\\(") |> str_remove("]"))) |> 
    separate(week, c("from", "to")) |> 
    mutate(across(where(is.character), as.integer),
           x = (from + to) / 2)
}
```

Calculate incidence from `start` by steps of `by` but sliding `start` by steps of 1
day:

```{r}
incidence <- function(start, by) {
  (start + 0:(by - 1)) |>
    map_dfr(~ incidence_from(.x, by)) |> 
    arrange(x) |> 
    na.exclude()
}
```

Calculate daily weekly incidences:

```{r}
w_incidence <- from <- df23 |> 
  pull(time) |> 
  min() |>
  incidence(7)
```

A function that plots the incidence `y` as a function of time `x`:

```{r}
plot_ts <- function(x, y, ...) {
  plotl(x, y, xaxt = "n", xlab = NA, ylab = "weekly incidence", ...)
  ats <- ymd(c(paste0("2023", sprintf("%02d", 1:12), "01"), "20240101"))
  axis(1, as.numeric(ats), month(ats, TRUE))
}
```

Plotting it:

```{r}
with(w_incidence, plot_ts(x, n, col = 4))
```


## Finite mixture model

Building the model:

```{r}
mclust_out <- df23 |>
  filter(time > 19490) |> 
  select(time) |> 
  Mclust(2) |> 
  extract2("parameters")

means <- mclust_out$mean
stdev <- sqrt(mclust_out$variance$sigmasq)

pnorm1 <- function(q) pnorm(q, means[1], stdev[1], FALSE)
pnorm2 <- function(q) pnorm(q, means[2], stdev[2])
```

Using the model to separate the two peaks

```{r}
two_peaks <- w_incidence |> 
  mutate(peak1 = pnorm1(x),
         peak2 = pnorm2(x),
         total = peak1 + peak2,
         peak1 = n * peak1 / total,
         peak2 = n * peak2 / total)

with(two_peaks, {
  plot_ts(x, peak1, col = 2)
  lines2(x, peak2, col = 4)
})

with(w_incidence, lines2(x, n, lty = 2))
```


## Age distributions

Computing the thresholds:

```{r}
th_p1 <- two_peaks |> 
  filter_out(peak2 > 2) |> 
  pull(x) |> 
  max()

th_p2 <- two_peaks |> 
  filter_out(peak1 > 2) |> 
  pull(x) |> 
  min()
```

Split the data set into peaks

```{r}
df23 <- mutate(df23, peak = ifelse(time < th_p1, "peak1",
                                   ifelse(time > th_p2, "peak2", "unknown")))
```

```{r}
df23 |> 
  group_by(peak) |> 
  tally()
```

```{r}
df23 |> 
  filter(peak == "peak1") |> 
  pull(age1) |> 
  density2() |> 
  plotd(xlim = c(0, 10), col = 2, xlab = "age (years)", ylab = "density")

df23 |> 
  filter(peak == "peak2") |> 
  pull(age1) |> 
  density2() |> 
  with(lines2(x, y, col = 4))

abline(v = 0:10, col = "grey")
```

```{r}
df23 |> 
  filter(peak == "peak1") |> 
  with(plot(admission_date, age1, ylim = c(0, 10)))
```

```{r}
predict2 <- function(object, newdata, ci = .95, ...) {
  if (! is.data.frame(newdata)) newdata <- as_tibble(newdata)
  bind_cols(newdata,
            as_tibble(predict(object, newdata, se = ci, type = "response", ...)))
} 
```

```{r}
tmp <- filter(df23, peak == "peak1")
mgcv::gam(age1 ~ s(time, k = 30), data = tmp) |> 
  predict2(list(time = seq(min(tmp$time), max(tmp$time), le = 512))) |> 
  mutate(across(time, as_date)) |> 
  with({
    plotl(time, fit, ylim = c(1, 3.5))
    lines2(time, ll)
    lines2(time, ul)
  })
```

```{r}
df23 |> 
  filter(peak == "peak1") |> 
  arrange(time) |> 
  slice(1:100) |> 
  summarise(time = mean(time), age = list(quantile(age1)))
```

```{r}
f <- function(x = 0, w = 100) {
  df23 |> 
    filter(peak == "peak1") |> 
    arrange(time) |> 
    slice(1:w + x) |> 
    summarise(time = mean(time),
              age = list(quantile(age1)))
}
```



```{r}
out <- map_dfr(1:(nrow(df23) - 100), f) |> 
  unnest_wider(age)
```


## Births

```{r}
hcm_births_perma <- readRDS2("linelisting/hcm_births_perma.rds")
hcm_births_register <- readRDS2("linelisting/hcm_births_register.rds")
```

```{r}
dim(hcm_births_perma)
dim(hcm_births_register)
```

```{r}
hcm_births_perma |> 
  group_by(dob) |> 
  tally() |> 
  with(plot(dob, n))
```

```{r}
hcm_births_register |> 
  group_by(dob) |> 
  tally() |> 
  with(plot(dob, n))
abline(v = 2014:2022)
```






